Unsupervised embedding of trajectories captures the latent structure of mobility

### 2020-10-01 paper/current/main.tex
intro edits

Here, we propose to apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which not only a meaningful \textit{functional distance} between places but also relationships based on culture, language, and even prestige along with geographic relationship can be probed. 
We leverage two massive datasets: a dataset of U.S. airport itinerary records and  scientific mobility captured in bibliometric records. In particular, we focus on scientific mobility throughout the paper due to its richness and importance. Scientific mobility, which is a central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, altbach2004globalization, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}, is not only an important topic on the science of science but also ideal for our study thanks to its richness and well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{auriol2010careers, clauset2015hierarchy, deville2014career}. 
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

For U.S. airport itinerary data, we use trajectories consists of the 828 unique identifier of U.S. airport in more than 300 million passenger flight itineraries between 1993 and 2020 from the Airline Origin and Destination Survey. Each trajectory includes origin, destination, and other itinerary details of passengers transported.
 For scientific mobility, we use trajectories of more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.
Each trajectory is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).

@@

To validate our approach, we evaluate the quality of vector representations with their performance in predicting real-world mobility flows using the gravity model framework~\autocite{zipf1946gravity, barbosa2018human}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, lewer2008immigrationgravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity} that models the \textit{expected flux}, $\hat{T}_{ij}$ between locations (organizations) based on their populations and distance:

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ is the population of locations $i$,  $f(r_{ij})$ is a decay function with respect to distance between locations, and $C$ is a constant estimated from data (see Data and Methods). For the U.S. airport itinerary, we use population $m_{i}$ as the total number of unique passenger on each airport and  for scientific mobility, we use the total number of mobile and non-mobile authors who affiliated with organization.
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of location $i$ and $j$ in the trajectory in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, is symmetric (\ie $\hat{T}_{ij }= \hat{T}_{ji})$, is proportional to the locations' population, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between locations or their functional distance in the vector space, which is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}. For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

>>>>

Here, we propose to apply a neural embedding framework to real-world mobility trajectories and demonstrate that it can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which we can not only derive a meaningful \textit{functional distance} between places but also probe relationships based on culture, language, and even prestige along with geographic relationship. 
We leverage two massive datasets: a dataset of U.S. passenger flight itinerary records, and a dataset of scientific mobility between organizations captured in bibliometric records. 
In particular, we focus on scientific mobility throughout the paper due to its richness and importance. Scientific mobility, which is a central driver of the globalized scientific enterprise~\autocite{czaika2018globalisation, altbach2004globalization, box2008competition} and strongly related to innovation~\autocite{braunerhjelm2020labor, kaiser2018innovation, armano2017innovation}, impact~\autocite{sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}, collaboration~\autocite{rodrigues2016mobility}, and the diffusion of knowledge~\autocite{braunerhjelm2020labor, azoulay2011diffusion, morgan2018prestige}, is not only an important topic on the science of science but also ideal for our study thanks to its richness and well-known structural properties such as the centrality of scientifically advanced countries and the strong prestige hierarchy~\autocite{auriol2010careers, clauset2015hierarchy, deville2014career}. 
In spite of its importance, understandings of scientific mobility have been limited by the sheer scope and complexity of the phenomenon~\autocite{robinson2019mobility, vannoorden2012mobility, deville2014career, ackers2008mobility, scott2015dynamics}, being further confounded by the diminishing role of geography in shaping the landscape of scientific mobility.

The flight itinerary data, from the Airline Origin and Destination Survey, consists of records of more than 300 million itineraries between 1993 and 2020 documenting domestic flights between 828 airports in the United States. 
A trajectory is constructed for each passenger flight itinerary, forming an ordered sequence of unique identifiers of the origin and destination airports. 
Although more complex, trajectories of scientific mobility are constructed in a similar fashion. 
Trajectories are constructed using more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.
Each trajectory is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).

@@

To validate our approach, we evaluate the quality of vector representations with their performance in predicting real-world mobility flows using the gravity model framework~\autocite{zipf1946gravity, barbosa2018human}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, lewer2008immigrationgravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity} that models the \textit{expected flux}, $\hat{T}_{ij}$ between locations (airports and organizations) based on their populations and distance:

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}

where $m_{i}$ is the population of locations $i$,  $f(r_{ij})$ is a decay function with respect to distance between locations, and $C$ is a constant estimated from data (see Data and Methods). 
For the flight itinerary data, we use population $m_{i}$ as the total number of unique passenger on each airport, and for scientific mobility, we use the mean annual number of of unique mobile and non-mobile authors who affiliated with each organization.
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of location $i$ and $j$ in the trajectory in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, is symmetric (\ie $\hat{T}_{ij }= \hat{T}_{ji})$), is proportional to the locations' population, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between locations or their functional distance in the vector space, which is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}. 
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).
